Vagueness, Noise, and Communication

### 2014-10-01 paper/04_discussion.tex
discussion section; oconnor

Finally, the replicator dynamics are also related to some forms of reinforcement learning (\cite{Borgers1997,Hopkins2005,Beggs2005}), and can thus be interpreted in cognitive terms.

\mytodo{JPC}{Discuss nature of noise}
\mytodo{JPC}{Compare with O'Connor}
\mytodo{JPC}{Compare with Franke2013}

>>>>

\subsection{Relation with previous accounts}

\paragraph{Generalized reinforcement.}
\citet{OConnor2013:The-Evolution-o} introduces a generalization of
Herrnstein reinforcement learning for sim-max games and showed that
this not only leads to vague signaling patterns, but can speed-up
evolution of efficient signaling strategies, especially in games with
many states.

Under plain Herrnstein reinforcement learning, sender and receiver
play the game repeatedly and adjust their dispositions to act after
each round of play, in such a way that the actual (non-negative)
payoff gained in the current interaction is added to the
non-normalized propensities for acting in exactly the way that they
acted in the current round of play. For sim-max games, this means that
when the sender chose $\messg$ in state $\state$, and this resulted in
some non-negative payoff (which is guaranteed by our choice of utility
function), the probability that the sender chooses $\messg$ again in
$\state$ is increased, but nothing else changes. In particular, the
speakers behavior in other choice points does not change. Generalized
reinforcement learning is different here. When the use of $\messg$ in
$\state$ gave positive payoff, then not only will its future use
probability be promoted at $\state$ but also at other states,
proportional to how similar these are to $\state$. Similar amendments
take care of the way that the receiver updates his choice
dispositions.

\citet{OConnor2013:The-Evolution-o} shows that this extension not only
leads to vague signaling of the appropriate kind, but also speeds up
learning in such a way that, especially for games with higher numbers
of states, higher levels of communicative success are reached in
shorter learning periods than is possible without stimulus
generalization. Technically, this result is partly due to the fact
that signalers make bigger changes to their behavioral strategies
after each round of play under generalized reinforcement than under
the plain variety. But that only explains the speed of adaptation, not
necessarily that generalization also leads to regularity and
communicative efficiency.

Diffusion of strategies in the \rdd can be conceived of as a form of
generalization as well, and works in large part quite analogous to
stimulus generalization in \citeauthor{OConnor2013:The-Evolution-o}'s
approach. This holds for the effect of diffusion and generalization,
but not necessarily for the way that the effect is achieved. We also
saw that diffusion in \rdd leads to more regular languages and
speedier convergence. However, the \rdd is a more abstract framework
than generalized reinforcement learning. The latter is foremost
motivated as a learning dynamic that has two players adapt their
individual strategies after each concrete round of play. In contrast,
the \rdd describes a more abstract, average dynamical change in
behavioral dispositions. Although the dynamics of (some forms of)
reinforcement learning mirror those of the replicator dynamics (at
some stage in time) (\cite{Borgers1997,Hopkins2005,Beggs2005}), this
does not mean that \emph{generalized} reinforcement learning, in which
stimulus generalization is best motivated at the level of a single
agent's dispositional generalization after one round of play, is also
directly a plausible high-level description of, say, generalized
learning in a population of agents. Seen in this light, generalized
reinforcement learning and the \rdd nicely complement each other, as
similarly-minded accounts operating at different levels of
abstraction.

\paragraph{Quantal response equilibria.}
\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in
which information processing limitations of signaling agents could
lead to vague strategies. The model that is most clearly related to
the present approach uses the notion of a quantal response, also known
as a logit response or a soft-max function
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,McKelveyPalfrey1995:Quantal-Respons,McKelveyPalfrey1998:Quantal-Respons,GoereeHolt2008:Quantal-Respons}. A
quantal response function is a paramterized generalization of the
classic best response function. For example, if $U \mycolon \Acts
\rightarrow \mathds{R}$ is the measure of expected utility over
choices $\Acts$ of an agent, then a best response function would have
the agent choose $\act$ only if $U(\act) = \max_{\act' \in \Acts}
U(a)$. A quantal response function rather assumes that agents would
choose $\act$ with a probability proportional to $\expo(\lambda \cdot
U(\act))$, where $\lambda$ is a rationality parameter. If $\lambda
\rightarrow \infty$ we retrieve the behavior of the best response
choice function, but if it is positive but finite, any choice $\act$
will receive a positive probability, but acts with higher expected
utility will be more likely. The underlying motivation for this choice
rule is the assumption that there is noise in the computation of
expected utilities and/or in maximization of expected
utilities. Consequently, choices with almost equal expected utilities
will be chosen with almost equal probability (for moderate values of
$\lambda$).

\citet{FrankeJager2010:Vagueness-Signa} show that quantal response
equilibria of sim-max games, i.e., pairs of sender and receiver
strategies such that the sender strategy is the quantal response to
the expected utilities under the receiver strategy and vice versa, can
show the desired marks of vague
signaling. Figure~\ref{fig:exampleQRE_stratsA} gives an example of a
quantal response equilibrium for a sim-max game, as used in our set-up
but with higher tolerance $\toler = 0.5$. Sender and receiver
strategies look very much like what evolves under \rdd with modest
values of perceptual imprecision.

\begin{figure}
  \centering

  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/exampleStratQRE_tolerance05.pdf}
    \caption{$\ns = 90$, $\lambda = 15$, $\toler = 0.5$}
    \label{fig:exampleQRE_stratsA}
  \end{subfigure}
  \hfill
  \begin{subfigure}[]{0.45\textwidth}
    \includegraphics[width=\textwidth]{plots/exampleStratQRE_tolerance005.pdf}
    \caption{$\ns = 90$, $\lambda = 15$, $\toler = 0.05$}
    \label{fig:exampleQRE_stratsB}
  \end{subfigure}

  \caption{Examples of vague quantal response equilibria.}
  \label{fig:exampleQREs}
\end{figure}

But not all quantal response equilibria are equally plausible
explanations of vague language use, and the ones that are not suggest
that it is less plausible to think of vagueness as arising because of
errors in the computation and maximization of expected utility, as the
quantal response approach assumes, than that it arises due to
confusion of similar states, as the \rdd proposes. To see what the
problem is, we can look at cases like given in
Figure~\ref{fig:exampleQRE_stratsB}, which is the quantal response
equilibrium for a game with lower tolerance $\toler = 0.05$. Unlike
what evolves under \rdd in this case, sender strategies have vague
boundaries also towards the end of the unit intervals. Technically,
this is because quantal responses equalize message use far away from
the ``prototypical'' interpretation, not just in-between categories,
so to speak. This, in turn, is because quantal responses introduce
noise into the decision making at the level of computing or maximizing
expected utility of choices.

That this is conceptually odd shows even more clearly in a case where
the state space is intuitively unbounded, as for instance for the
property ``tall''. If the usual interpretation of a ``tall man'' peaks
at around, say, 195cm then when meeting a giant of $n$ meters speakers
would, according to the quantal response approach, be ever more
inclined to describe the giant indifferently as either ``tall'' or
``short'' the larger $n$ gets. This is because, as the distance from
the prototype increases for larger $n$, the difference between the
expected utilities of saying ``short'' or ``tall'' will converge to
zero. Whence that the quantal response approach would predict that
speakers would grow indifferent between choice of antonyms as $n$
grows, which seems weird.

Admittedly, this argument hinges on the choice of utility
function. Still, to the extent that the chosen lower bounded utility
functions are reasonable ---and we think they are very reasonable---,
the case suggests that quantal responses are not a good model,
intuitively speaking, for why linguistic categories are
vague. Vagueness is more plausibly an effect of perceptual confusion
of similar states, than of computational errors in maximizing expected
utility.
